|

### 2020-07-28 article_format/article_format.Rmd
Changed the formating, made some minor edits

---
title: \singlespacing A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
author: Henrik Kenneth Andersen
abstract: "This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model. \\par  \\textbf{Keywords:} Fixed effects, structural equation modeling, lavaan, R, panel analysis" 
note: "Preprint version, please do not cite without permission."
date: "`r Sys.Date()`"
geometry: margin=1in
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
urlcolor: blue
link-citations: yes
bibliography: references.bib   
fontsize: 12pt
---

@@

# Script hooks for chunk outputs
source("scripthooks.R")
```

@@

@@

The following article outlines the basic idea of FE regression, the particularities of FE in SEM, and shows its implementation in `lavaan`. Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. One of the main strengths of panel SEM compared to the more traditional methods of panel analysis is its flexibility. Therefore, a number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).

@@

To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) of the article when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align}
where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$. $\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.

@@

Consistency of the following models requires the assumption of strict exogeneity, although what constitutes strict exogeneity differs between the random and fixed effects setups. Each assumption will be discussed shortly. Apart from that, we typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019] about this model:

@@

@@

The second part is the potentially controversial assumption: it states that the individual effects are uncorrelated with the independent variables. We can use an example to show why this is often controversial: If we are interested in the question of whether married men earn more than unmarried men, then the second part of the strict exogeneity assumption means that a man's marriage status would have to be uncorrelated with all the time-invariant characteristics that could potentially make that man an attractive marriage candidate in the first place; e.g., looks, personality, family's status, profession, etc. [@Bruederl2015].^[Assuming, for the sake of argument, that these characteristics are constant over time.]

From what we have discussed so far, the $T \times T$ covariance matrix of the errors $\bm{\Omega}_{i} = \E[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]$ can be constructed. However, the standard random effects model adds the additional assumptions 
\begin{alignat}{3}
\E[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
\E[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
\end{alignat}
i.e., the idiosyncratic errors are conditionally homoscedastic and serially uncorrelated and 
\begin{alignat}{2}
\E[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \E[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
\end{alignat}
i.e., the individual effects are conditionally homoscedastic (they are necessarily serially correlated as long as $\sigma_{\alpha}^{2} > 0$) we arrive at the typical random effects structure of the $NT \times NT$ matrix $\bm{\Omega}$:

@@

In practice, however, computational problems can arise with large cross-sectional samples, where it can become difficult to invert the $\hat{\bm{\Omega}}$ matrix. One solution is to use 'partial-demeaning' to transform the data before performing simple OLS:
\begin{align}
(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
\end{align} 
where $\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}$, and $\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}$, $\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}$, $\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}$ and $\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}$ [@R-plm_a].

The RE model can also be estimated in the maximum likelihood framework, where in the associated literature panel models are generally referred to as either mixed models, hierarchical models or longitudinal models. The typical RE model discussed here is the equivalent to a mixed model with random intercepts and fixed slopes [@R-plm_a]. Under the assumption of normality, along with homoscedasticity and serially uncorrelated errors, the maximum likelihood estimator is the same as the OLS estimator. For more on the topic of mixed models, see for example @R-lme4, @Bates2010.

@@

And in fact this is exactly how a latent variable is used to account for time-invariant unobserved heterogeneity. The dependent variable at each timepoint is regressed onto the latent variable, see Figure \ref{fig:fesem}. Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time. It also means that the estimated variance of the latent variable is equal to the \textit{average covariance between the wide-format columns of the dependent variable over time}. If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\alpha$, regardless of the time distance. Referring back to the random effects structure of $\bm{\Omega}_{i}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$. And, as we know, the average of something that does not change is that thing itself. I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{i}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.

To show this, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
\begin{align}
\bm{A}\bm{x} & = \bm{b} \\
\begin{pmatrix}
1 & 1 & 0 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 1 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 0 & 1
\end{pmatrix}
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix} & = 
\begin{pmatrix}
\Var(y_{1}) \\
\Cov(y_{2},y_{1}) \\
\Cov(y_{3},y_{1}) \\
\Var(y_{2}) \\
\Cov(y_{3},y_{2}) \\
\Var(y_{3})
\end{pmatrix}
\end{align}
where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation to show
\begin{align}
\bm{A}\bm{x} & = \bm{b} \\
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
\hat{\bm{x}} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
\begin{pmatrix}
\hat{\psi} \\
\hat{\phi}_{1} \\
\hat{\phi}_{2} \\
\hat{\phi}_{3}
\end{pmatrix} & = 
\begin{pmatrix}
\frac{1}{3}\Cov(y_{2},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{2}) \\
\Var(y_{1}) - \hat{\psi} \\
\Var(y_{2}) - \hat{\psi} \\
\Var(y_{3}) - \hat{\psi}
\end{pmatrix}.
\end{align}
This shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time. Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.

@@

What essentially differentiates an FE from a random effects (RE) model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010]. The FE model assumes that $\E[\alpha x_{t}] \ne 0$. As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$, will be biased. Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
\begin{align}
\bm{y}^{+}\bm{y}^{+ \intercal} & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
 & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
\end{align}
In the case of an FE model, $\bm{\Psi}$ will reflect our belief that the individual effects are correlated with the model covariates, here again for demonstration the three-wave model:
\begin{align}
\bm{\Psi} & = \E
\begin{blockarray}{cccccccc}
 & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
 \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
 \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
 x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
 x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
 x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
 \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
 \end{block}
\end{blockarray}.
\end{align}
Knowing this, we can work out the equation for the coefficient of interest, $\beta$. For the sake of simplicity, assume here and throughout mean-centered variables: 
\begin{align}
\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
 & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
 & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
 & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
\end{align}
This should make intuitive sense. From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual. For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})$. The rest of the model-implied covariance matrix results from $\bm{y}^{+}\bm{y}^{+ \intercal}$.

@@

```{r echo=FALSE}
df <- readRDS("longData.Rda")
dfw <- readRDS("wideData.Rda")
```

@@

```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
saveRDS(fe_sem, "fe_sem.Rda")
saveRDS(fe_sem.fit, "fe_sem.fit.Rda")
```

@@

Fixed effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001]. This article provides a focused look at the implementation of the basic model, as well as common extensions using the `lavaan` package in `R`.

@@

The most basic FE-SEM is furthermore the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014]. For this reason it is all the more important for researchers to have a good grasp on the method of applying FE-SEM. This article is meant to serve as a consolidated resource for researchers looking for concrete advise on specifying FE and more general panel models in SEM.

# References {-}

<div id="refs"></div>

>>>>

---
title: |
  A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
#type: ARTICLE TEMPLATE
author:
  - name: Henrik Kenneth Andersen
    affil: a
    email: henrik.andersen@soziologie.tu-chemnitz.de
affiliation:
  - num: a
    address: |
      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
bibliography: references2.bib
# appendix: appendix.tex
abstract: |
  This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extensions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model.
keywords: |
  Fixed effects, structural equation modeling, lavaan, R, panel analysis
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{booktabs}
  \usepackage{bm}
  \usepackage{mathtools}
  \usepackage{amssymb}
  \usepackage{amsmath}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \usepackage[nofiglist]{endfloat}
  \usepackage{blkarray}
  \usepackage{setspace}
  \usepackage{etoolbox}
  \DeclareMathOperator{\E}{\mathbb{E}}
  \DeclareMathOperator{\Var}{\mathrm{Var}}
  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
  \DeclareMathOperator*{\argmax}{arg\,max}
  \DeclareMathOperator*{\argmin}{arg\,min}
  \mathtoolsset{showonlyrefs}
  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
  \def\tightlist{}
  \pagenumbering{gobble}
output: rticles::tf_article
---

@@

# Script hooks for chunk outputs
source("../scripthooks.R")
```

@@

\doublespacing

@@

The following article outlines the basic idea of panel regression, the particularities of panel regression SEM, and shows its implementation in `lavaan`. The focus on FE panel regression stems from the fact that the assumptions needed to justify other common models, e.g., random effects, are often implausible. As we will see, the practical differences between the two are, in any event, very small. Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. One of the main strengths of panel SEM compared to the more traditional methods of panel analysis is its flexibility. Therefore, a number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).

@@

To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align}
where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$. $\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.

@@

Consistency of the following models requires the assumption of strict exogeneity, although what constitutes strict exogeneity differs between the random and fixed effects setups. Each assumption will be discussed shortly. Apart from that, we typically make the following assumptions about this model [see, e.g., @Wooldridge2002; @Schmidheiny2019]:

@@

Somewhat unintuitively, the modern approach to FE regression treats the unobserved effect $\alpha_{i}$ as a \textit{random} variable, rather than a fixed parameter to be estimated for each of the cross sectional units [@Wooldridge2002]. And in fact, this view carries over to the way in which FE regression is applied in SEM. So, in order to help facilitate the discussion on the application of FE-SEM later on, it makes sense to spend some time discussing the basic idea of random effects regression (again, as we will see, the practical difference between the two models is very small).

@@

The second part is the potentially controversial assumption: it states that the individual effects are uncorrelated with the independent variables. We can use an intuitive concrete example to show why this is often controversial: If we are interested in the question of whether married men earn more than unmarried men, then the second part of the strict exogeneity assumption means that a man's marriage status would have to be uncorrelated with all the time-invariant characteristics that could potentially make that man an attractive marriage candidate in the first place; e.g., looks, personality, family's status, profession, etc. [@Bruederl2015].^[Assuming, for the sake of argument, that these characteristics are constant over time.]

From what we have discussed so far, the $T \times T$ covariance matrix of the errors $\bm{\Omega}_{i} = \E[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]$ can be constructed. However, the standard random effects model adds the additional assumptions 
\begin{alignat}{3}
\E[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
\E[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
\end{alignat}
i.e., the idiosyncratic errors are conditionally homoscedastic and serially uncorrelated and 
\begin{alignat}{2}
\E[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \E[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
\end{alignat}
i.e., the individual effects are conditionally homoscedastic (they are necessarily serially correlated as long as $\sigma_{\alpha}^{2} > 0$). From that, we arrive at the typical random effects structure of the $NT \times NT$ matrix $\bm{\Omega}$:

@@

In practice, however, computational problems can arise with large cross-sectional samples, where it can become difficult to invert the $\hat{\bm{\Omega}}$ matrix. One solution is to use 'partial-demeaning' to transform the data before performing simple POLS:
\begin{align}
(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
\end{align} 
where $\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}$, and $\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}$, $\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}$, $\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}$ and $\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}$ [@R-plm_a].

The RE model can also be estimated in the maximum likelihood framework, where in the associated literature on panel models are generally referred to as either mixed models, hierarchical models or longitudinal models. The typical RE model discussed here is the equivalent to a mixed model with random intercepts and fixed slopes [@R-plm_a]. Under the assumption of normality, along with homoscedasticity and serially uncorrelated errors, the maximum likelihood estimator is the same as the OLS estimator. For more on the topic of mixed models, see for example @R-lme4, @Bates2010.

@@

And in fact this is exactly how a latent variable is used to account for time-invariant unobserved heterogeneity. The dependent variable at each timepoint is regressed onto the latent variable, see Figure \ref{fig:fesem}. Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time.^[In fact, the initial FE-SEM setup shown in the main article mimics the POLS methods described above in that it assumes constant effects and error variances over time. These assumptions can be loosened and tested, as will be shown in the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf). For now, for the sake of simplicity and comparability, we retain the assumptions associated with the 'pooled' models for the most part.] It also means that the estimated variance of the latent variable is equal to the \textit{average covariance between the wide-format columns of the dependent variable over time}. If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\Var(\alpha)$, regardless of the time distance. Referring back to the random effects structure of $\bm{\Omega}_{i}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$. And, as we know, the average of something that does not change is that thing itself. I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{i}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.

To elaborate on this concept some more, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
\begin{align}
\bm{A}\bm{x} & = \bm{b} \\
\begin{pmatrix}
1 & 1 & 0 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 1 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 0 & 1
\end{pmatrix}
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix} & = 
\begin{pmatrix}
\Var(y_{1}) \\
\Cov(y_{2},y_{1}) \\
\Cov(y_{3},y_{1}) \\
\Var(y_{2}) \\
\Cov(y_{3},y_{2}) \\
\Var(y_{3})
\end{pmatrix}
\end{align}
where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation to show
\begin{align}
\bm{A}\bm{x} & = \bm{b} \\
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
\hat{\bm{x}} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
\begin{pmatrix}
\hat{\psi} \\
\hat{\phi}_{1} \\
\hat{\phi}_{2} \\
\hat{\phi}_{3}
\end{pmatrix} & = 
\begin{pmatrix}
\frac{1}{3}\Cov(y_{2},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{2}) \\
\Var(y_{1}) - \hat{\psi} \\
\Var(y_{2}) - \hat{\psi} \\
\Var(y_{3}) - \hat{\psi}
\end{pmatrix}.
\end{align}
So if, in fact the covariance between any two wide-format columns of $y$ is $\Cov(y_{t},y_{s}) = \Var(\alpha), \ \forall \ s \ne t$, then $\hat{\psi} = \frac{1}{3}\Var(\alpha) + \frac{1}{3}\Var(\alpha) + \frac{1}{3}\Var(\alpha) = \frac{3 \Var(\alpha)}{3} = \Var(\alpha)$. This shows that if our assumption about the underlying DGP is correct, i.e., $y_{it} = \alpha_{i} + \varepsilon_{it}$ and $\Cov(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s$, then the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time. Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.

@@

What essentially differentiates an FE from an RE model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010]. The FE model assumes that $\E[\alpha x_{t}] \ne 0$. As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$, will be biased. Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
\begin{align}
\bm{y}^{+}\bm{y}^{+ \intercal} & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
 & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
\end{align}
In the case of an FE model, $\bm{\Psi}$ will reflect our belief that the individual effects are correlated with the model covariates, here again for demonstration the three-wave model:
\begin{align}
\bm{\Psi} & = \E
\begin{blockarray}{cccccccc}
 & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
 \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
 \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
 x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
 x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
 x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
 \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
 \end{block}
\end{blockarray}.
\end{align}
Knowing this, we can work out the equation for the coefficient of interest, $\beta$. For the sake of simplicity, assume here and throughout mean-centered variables: 
\begin{align}
\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
 & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
 & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
 & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
\end{align}
This should make intuitive sense. From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual. For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})$. The rest of the model-implied covariance matrix results from $\bm{y}^{+}\bm{y}^{+ \intercal}$.

@@

```{r echo=FALSE}
df <- readRDS("../longData.Rda")
dfw <- readRDS("../wideData.Rda")
```

@@

```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
saveRDS(fe_sem, "../fe_sem.Rda")
saveRDS(fe_sem.fit, "../fe_sem.fit.Rda")
```

@@

Fixed effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001]. This article provides a focused look at the implementation of the basic model using the `lavaan` package in `R`. The [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) further discuss common extensions and some tools for evaluating and loosening model assumptions.

@@

The most basic FE-SEM is furthermore the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014]. For this reason it is all the more important for researchers to have a good grasp on the method of applying panel regression in SEM, and understanding the intuition of controlling for time-invariant confounders. This article is meant to serve as a consolidated resource for researchers looking for concrete advise on specifying FE and more general panel models in SEM.

# References
